Columnar VCF for Biobank scale datasets using Zarr

### 2024-04-11 paper.tex

\begin{figure}
\includegraphics{figures/subset-matrix-compute}
\caption{Compute performance on subsets of the matrix.
Total CPU time required to run the af-dist calculation for
a subset of 10000 variants $\times$ 10 samples from the middle of the matrix
for the data in Figure~\ref{fig-data-storage}.
The \texttt{bcftools} and \texttt{bcftools} pipelines involve
multiple commands required to correctly calculate the AF INFO tag
required by \texttt{bcftools +af-dist}. See the Methods for full details
on the steps performed.
\label{fig-subset-matrix-compute}}
\end{figure}

We illustrate the cost of row-wise encoding in
Figure~\ref{fig-subset-matrix-compute}, where we run the af-dist calculation
on a small fixed-size subset of the genotype matrices of
Figure~\ref{fig-data-storage}. The two-dimensional chunking of Zarr
means that this small subset of the overall matrix can be efficiently
extracted, and therefore the execution time depends very weakly on 
the overall dataset size, with the overall computation requiring around
1 second for 1 million samples. CPU time, for this simple calculation
on a fixed size sub-matrix scales, with the dataset size.
Figure~\ref{fig-subset-matrix-compute-supplemental} shows performance
for the same operation when selecting half of the samples in the 
dataset.

>>>>

\begin{figure}
\includegraphics{figures/subset-matrix-compute}
\caption{Compute performance on subsets of the matrix.
Total CPU time required to run the af-dist calculation for
a subset of 10000 variants $\times$ 10 samples from the middle of the matrix
for the data in Figure~\ref{fig-data-storage}.
The \texttt{genozip} and \texttt{bcftools} pipelines involve
multiple commands required to correctly calculate the AF INFO field
required by \texttt{bcftools +af-dist}. See the Methods for full details
on the steps performed.
\label{fig-subset-matrix-compute}}
\end{figure}

We illustrate the cost of row-wise encoding in
Figure~\ref{fig-subset-matrix-compute}, where we run the af-dist calculation
on a small fixed-size subset of the genotype matrices of
Figure~\ref{fig-data-storage}. The two-dimensional chunking of Zarr
means that this small subset of the overall matrix can be efficiently
extracted, and therefore the execution time depends very weakly on 
the overall dataset size, with the overall computation requiring around
1 second for 1 million samples. Because of their 
row-wise encoding for all the other methods CPU time
scales with the number of samples.
Figure~\ref{fig-subset-matrix-compute-supplemental} shows performance
for the same operation when selecting half of the samples in the 
dataset.
